---
title: "Applied Controlled Comparisons"
subtitle: POSC 3410  -- Quantitative Methods in Political Science
author: Steven V. Miller
institute: Department of Political Science
titlegraphic: /Dropbox/teaching/clemson-academic.png
date: 
output:
 beamer_presentation:
    template: ~/Dropbox/miscelanea/svm-r-markdown-templates/svm-latex-beamer.tex
    latex_engine: xelatex
    dev: cairo_pdf
    fig_caption: true
    slide_level: 3
make149: true
mainfont: "Open Sans"
titlefont: "Titillium Web"
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
---

```{r setup, include=FALSE, cache=F, message=F, warning=F, results="hide"}
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_chunk$set(fig.path='figs/')
knitr::opts_chunk$set(cache.path='cache/')

knitr::opts_chunk$set(
                  fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      }
                  )
```

```{r loadpackages, include=FALSE, cache=FALSE}
library(tidyverse)
library(stevemisc)
#library(post8000r)
library(qs)
library(knitr)
library(kableExtra)

skable <- function(data, caption, aligns,  colnames) {
  require(knitr)
  require(kableExtra)
  knitr::kable(data, caption = caption, col.names = if(missing(colnames)) { names(data)} else { colnames }, format = "latex", booktabs = TRUE, longtable =T, linesep = "", align = if(missing(aligns)) {} else { aligns }) %>%
    kable_styling(full_width = F, 
                  latex_options = c("repeat_header", "hold_position")) %>%
    row_spec(0, bold= TRUE)
}
```

```{r loaddata, include=FALSE, cache=TRUE}
gunlaw <- qread("/home/steve/Dropbox/data/gss/GSS_spss-2018/gss7218.qs") %>% 
  select(year, gunlaw, partyid, sex) %>% 
  filter(year == 2018) %>%
  mutate(gender = case_when(
    sex == 2 ~ "Female",
    sex == 1 ~ "Male"
  ),
  gunlaw = carr(gunlaw, "1=1;2=0"),
  gunlawcat = ifelse(gunlaw == 1, "Favor", "Oppose"),
         pidcat = case_when(
           partyid %in% c(0,1, 2) ~ "Democrat/Lean Democrat",
           partyid == 3 ~ "Independent",
           partyid %in% c(4,5,6) ~ "Republican/Lean Republican"
         ))

abortimport <- haven::read_dta("/home/steve/Dropbox/data/anes/2020-ets/anes_pilot_2020ets_dta.dta") %>%
  select(abort1, abort_imp, pid7, vote16) %>%
  mutate(abort_always = carr(abort1, "9=NA;1:3=0;4=1"),
         abort_never = carr(abort1, "9=NA;1=1;2:4=0"),
         abort_impdum = carr(abort_imp, "9=NA;1:2=1;3:5=0"),
         pidcat = case_when(
           pid7 %in% c(1,2,3) ~ "Republican/Lean Republican",
           pid7 == 4 ~ "Independent",
           pid7 %in% c(5,6,7) ~ "Democrat/Lean Democrat"
         ),
         vote16cat = case_when(
           vote16 == 1 ~ "Donald Trump",
           vote16 == 2 ~ "Hillary Clinton",
           vote16 == 3 ~ "Someone Else",
           vote16 == 4 ~ "Did Not Vote"
         )) %>% haven::zap_labels() %>%
  mutate(votetrump = ifelse(vote16cat == "Donald Trump", 1, 0))

```

# Introduction
### Goals for Today

- *Give applied examples of controlled comparisons and controlled relationships.*
- *Show how to do some of these with R code.*
- *Discuss controlled effects and partial effects.*

### Three Types of Controlled Relationships

1. Spurious relationship
2. Additive relationship
3. Interactive relationship

Notice our discussion here is *not* bivariate.

- We need to consider "control" for potential confounders in a bivariate relationship.

# Applied Controlled Comparisons
### Applied Controlled Comparisons

We can summarize these controlled relationships with:

1. Cross-tabulation
2. Mean comparison analysis

We will also use actual data.



### Example Data

We'll be going over two toy examples, largely patterned off Pollock:

1. Attitudes about gun control as a function of partisanship and gender.
2. Vote choice in 2016 as a function of abortion attitudes and issue importance.

Data come from General Social Survey (2018) and ANES (2020, ETS), respectively.

### R Packages

```r
library(tidyverse) # for most things
library(qs) # for loading my version of the GSS data
library(knitr) # for tables
library(kableExtra) # for prettier tables
```

### 

\scriptsize
```{r, eval=F}
# load data, assuming this qs object
# see: http://svmiller.com/blog/2020/02/comparing-qs-fst-rds-for-bigger-datasets/
gunlaw <- qread("~/Dropbox/data/gss/GSS_spss-2018/gss7218.qs") %>% 
  # select just what we want
  select(year, gunlaw, partyid, sex) %>% 
  filter(year == 2018) %>%
  # recode sex to gender category
  mutate(gender = case_when(
    sex == 2 ~ "Female",
    sex == 1 ~ "Male"
  ),
  # recode data so that 1 = favor
  gunlaw = carr(gunlaw, "1=1;2=0"),
  # Group the leaners with the partisans
  gunlawcat = ifelse(gunlaw == 1, "Favor", "Oppose"),
         pidcat = case_when(
           partyid %in% c(0,1, 2) ~ "Democrat/Lean Democrat",
           partyid == 3 ~ "Independent",
           partyid %in% c(4,5,6) ~ "Republican/Lean Republican"
         ))
```
\normalsize

###

\scriptsize

```{r, eval=F}
# load data, from here...
# https://electionstudies.org/data-center/2020-exploratory-testing-survey/
haven::read_dta("~/Dropbox/data/anes/2020-ets/anes_pilot_2020ets_dta.dta") %>%
  # select just what we want
  select(abort1, abort_imp, pid7, vote16) %>%
  # recode that we get the "never permit abortion" people as 1s
  # and whether the issue is extremely or very important
  # also lump leaners with the partisans and create vote choice variables
  mutate(abort_never = carr(abort1, "9=NA;1=1;2:4=0"),
         abort_impdum = carr(abort_imp, "9=NA;1:2=1;3:5=0"),
         pidcat = case_when(
           pid7 %in% c(1,2,3) ~ "Republican/Lean Republican",
           pid7 == 4 ~ "Independent",
           pid7 %in% c(5,6,7) ~ "Democrat/Lean Democrat"
         ),
         vote16cat = case_when(
           vote16 == 1 ~ "Donald Trump",
           vote16 == 2 ~ "Hillary Clinton",
           vote16 == 3 ~ "Someone Else",
           vote16 == 4 ~ "Did Not Vote"
         )) %>% haven::zap_labels() %>%
  mutate(votetrump = ifelse(vote16cat == "Donald Trump", 1, 0)) -> abortimport
```
\normalsize

### Partisanship and Gun Control (GSS, 2018)

```{r, echo=F, message=F, warning=F, results="asis"}

gunlaw %>%
  filter(pidcat != "Independent") %>%
  select(pidcat, gunlawcat) %>% na.omit %>%
  group_by(pidcat, gunlawcat) %>%
  tally() %>%
  spread(pidcat, n) %>%
  rowwise() %>%
  mutate(total = rowSums(across(where(is.numeric)))) %>%
  ungroup() %>%
  mutate_if(is.numeric,~(./sum(.))) %>%
  mutate_if(is.numeric, ~(paste0(mround2(.), "%"))) %>%
  bind_rows(., tibble(gunlawcat = "Total"))  %>%
  replace(is.na(.), "100%") %>%
  kable(., format="latex",
        caption="The Relationship Between Partisanship and Support for Gun Control",
         col.names = c("Opinion on Gun Permits", "D/Lean D", "R/Lean R", "Total"),
         align = c("l","c","c","c","c","c","c"), booktabs=TRUE, longtable=TRUE) %>%
  kable_styling(font_size = 9) %>%
  # add_header_above(c(" ", "Women" = 3, "Men"= 3), italic = TRUE) %>%
  row_spec(2, hline_after = T) %>%
  row_spec(0, bold= TRUE) %>%
  row_spec(3, italic = TRUE) %>%
  footnote(general = "Data: General Social Survey (2018)",
           footnote_as_chunk = TRUE)
  

```

### Zero-order Relationship

This table is a zero-order relationship.

- The effect of *x* on *y* **not** controlling for *z*.

In 2018, the zero-order effect of partisanship on support for gun control was about -22.87 percentage points.


### Partisanship and Gun Control (GSS, 2018)

```{r, echo=F, message=F, warning=F, results="asis"}

gunlaw %>%
  filter(pidcat != "Independent") %>%
  filter(gender == "Female") %>%
  select(pidcat, gender, gunlawcat) %>% na.omit %>%
  group_by(pidcat, gunlawcat) %>%
  tally() %>%
  spread(pidcat, n) %>%
  rowwise() %>%
  mutate(total = rowSums(across(where(is.numeric)))) %>%
  ungroup() %>%
  mutate_if(is.numeric,~(./sum(.))) -> women


gunlaw %>%
  filter(pidcat != "Independent") %>%
  filter(gender == "Male") %>%
  select(pidcat, gender, gunlawcat) %>% na.omit %>%
  group_by(pidcat, gunlawcat) %>%
  tally() %>%
  spread(pidcat, n) %>%
  rowwise() %>%
  mutate(total = rowSums(across(where(is.numeric)))) %>%
  ungroup() %>%
  mutate_if(is.numeric,~(./sum(.))) -> men

# men %>% 
#   rename(a = 2, b=3, c=4) -> men


bind_cols(women, men %>% select(-gunlawcat)) %>%
  mutate_if(is.numeric, ~(paste0(mround2(.), "%"))) %>%
  bind_rows(., tibble(gunlawcat = "Total")) %>%
  replace(is.na(.), "100%") %>% 
  kable(., format="latex",
        caption="Partisanship and Support for Gun Control, Controlling for Gender",
         col.names = c("Opinion on Gun Permits", "D/Lean D", "R/Lean R", "Total", "D/Lean D", "R/Lean R", "Total"),
         align = c("l","c","c","c","c","c","c"), booktabs=TRUE, longtable=TRUE) %>%
  kable_styling(font_size = 9) %>%
  add_header_above(c(" ", "Women" = 3, "Men"= 3), italic = TRUE) %>%
 row_spec(2, hline_after = T) %>%
    row_spec(3, italic = TRUE) %>%
  row_spec(0, bold= TRUE) %>%
  footnote(general = "Data: General Social Survey (2018)",
           footnote_as_chunk = TRUE)

```

## Controlled and Partial Effects
### Controlled Comparison Table

That was a **controlled comparison table**.

- It shows the relationship between *x* and *y* for each (truncated) value of *z*.

These tables reveal two types of effects.

1. Controlled effect
2. Partial effect

### Controlled Effect

A **controlled effect** is a relationship between *x* and *y* within one value of *z*.

- It could also be between *z* and *y* within one value of *x*.

We obtain the controlled effect of partisanship for both women and men.

- 85.89% of female Dems/Dem leaners favor these permits to 64.91% of Rs/R leaners.
    - The controlled effect is -20.98 percentage points.
- 76.87% of male Dems/Dem leaners favor these permits to 66.05% of Rs/R leaners.
    - The controlled effect is -23.81 percentage points.
    
### Partial Effect

We summarize controlled effects as **partial effect**.

- Summarizes a relationship between two variables taking into account rival variables.

It's tempting, but *don't average the two controlled effects*.

- Doing so assumes the samples are equal.

Instead: weight the controlled effect by percentage of the sample.

- Then: add those.


### 

\footnotesize
```{r}
gunlaw %>%
  select(gunlawcat, gender, pidcat) %>%
  na.omit %>%
  group_by(gender) %>% tally() %>%
  mutate(prop = n/sum(n),
         contreff = c(-20.98, -23.81),
         product = prop*contreff,
         parteff = sum(product))
```
\normalsize

The partial effect of partisanship on gun control opinions is about -22.2 percentage points.


### Partial Effect

What is the partial effect of *gender* on gun control attitudes?

- Sounds weird to ask. No one "increases" in gender.

Follow the **rule of direction for nominal relationships**.

- Treat the left-most column as the base category (here: women).

### Partial Effect

Controlled effects:

- 85.89% of female Dems/leaners favor these permits to 76.87% of male Dems/leaners (-9.02%).
- 64.91% of female Rs/leaners favor permits to 53.06% of male Republicans (-11.85%).

Weight the controlled effects (*by partisanship*) to get a partial effect.

<!-- - 10.7$*$(.570) + 13.8$*$(.430) = 12.03 -->

### 

\footnotesize
```{r}
gunlaw %>%
  select(gunlawcat, gender, pidcat) %>%
  # simple example excludes the indies
  filter(pidcat != "Independent") %>%
  na.omit %>%
  group_by(pidcat) %>% tally() %>%
  mutate(prop = n/sum(n),
         contreff = c(-9.02, -11.85),
         product = prop*contreff,
         parteff = sum(product))
```

\normalsize

The partial effect of gender on gun control opinions (controlling for partisanship) is about -10.2 percentage points.

## Identifying the Pattern
### Identifying the Pattern

Ask the following three questions for relationships among *x*, *y*, and *z*.

1. Does a relationship exist between *x* and *y* in at least one value of *z*?
2. Is the tendency (i.e. positive or negative) the same at all values of *z*?
3. Is the magnitude effect the same or close to it in all values of *z*?

### Identifying the Pattern

If the answer to the first question is no, you can stop there.

- It's a spurious relationship.

If the answer to the second question is no, you can stop there.

- There's an interaction effect.

If the answer to the third question is no, there's an interaction.

- If "yes", it's an additive relationship.

###

```{r partisanship-gun-control-gender, eval=T, echo=F, fig.width=14, fig.height=8.5, warning=F, message=F}
gunlaw %>%
  group_by(pidcat, gender) %>%
  summarize(mean = mean(gunlaw, na.rm=T)) %>%
  na.omit %>%
  filter(pidcat != "Independent") %>%
  ggplot(.,aes(x=pidcat, y = mean,
               group=gender,color=gender,
               linetype=gender, shape=gender)) + 
  theme_steve_web() +
  geom_point(size=4) + geom_line(size=1.1) +
  scale_color_brewer(palette="Paired") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "A Line Chart of Partisanship and Support for Gun Control, Controlling for Gender",
       subtitle = "The lines are effectively the same, implying an additive relationship. The distance between two points on the same part of the x-axis is the controlled effect of gender.",
       x = "",
       y = "Percentage Supporting a Police Permit Before Obtaining a Gun",
       linetype = "", color = "", shape="",
       caption = "Data: General Social Survey")
```


### Abortion Opinions, Issue Salience, and Vote Choice (ANES, 2020)

```{r, echo=F, message=F, warning=F, results="asis"}

abortimport %>%
  filter(abort_impdum == 0) %>%
  select(abort_impdum, votetrump, abort_never) %>%
  na.omit %>%
  group_by(abort_never, votetrump) %>%
  tally() %>%
  spread(abort_never, n) %>%
  rowwise() %>%
  mutate(total = rowSums(across(where(is.numeric)))) %>%
  ungroup()  %>%
  mutate_if(is.numeric,~(./sum(.))) -> abort_notimp

abortimport %>%
  filter(abort_impdum == 1) %>%
  select(abort_impdum, votetrump, abort_never) %>%
  na.omit %>%
  group_by(abort_never, votetrump) %>%
  tally() %>%
  spread(abort_never, n) %>%
  rowwise() %>%
  mutate(total = rowSums(across(where(is.numeric)))) %>%
  ungroup()  %>%
  mutate_if(is.numeric,~(./sum(.))) -> abort_veryimp



bind_cols(abort_notimp, abort_veryimp %>% select(-votetrump)) %>%
  arrange(-votetrump) %>%
  mutate(votetrump = c("Yes", "No")) %>%
  mutate_if(is.numeric, ~(paste0(mround2(.), "%"))) %>%
  bind_rows(., tibble(votetrump = "Total")) %>%
  replace(is.na(.), "100%") %>%
  kable(., format="latex",
        caption="Abortion Permissibility and the Trump Vote, Controlling for Issue Importance",
         col.names = c("Voted for Trump?", "Permit", "Never Permit", "Total",
                       "Permit", "Never Permit", "Total"),
         align = c("l","c","c","c","c","c","c"), booktabs=TRUE, longtable=TRUE) %>%
  kable_styling(font_size = 9) %>%
  add_header_above(c(" ", "Not Very Important" = 3, "Very Important"= 3), italic = TRUE) %>%
 row_spec(2, hline_after = T) %>%
    row_spec(3, italic = TRUE) %>%
  row_spec(0, bold= TRUE) %>%
  footnote(general = "Data: American National Election Studies (2020 Exploratory Testing Survey)",
           footnote_as_chunk = TRUE)


```


### Abortion Opinions, Salience, and Vote Choice

The controlled effect of abortion opinion:

- 33.33 -> 42.37 for "not very important" = `r abs(33.33 - 42.37)`
- 33.48 -> 62.13 for "very important" = `r abs(33.48 - 62.13)`

Controlled effect of issue importance:

- 33.33 -> 33.48 for "permit" = `r abs(33.33 - 33.48)`
- 42.37 -> 62.13 for "never permit" = `r abs(42.37 - 62.13)`
<!-- - Partial effect: -18.8 $*$ (.416) + 11 $*$ (.584) = -1.39 -->

Something already looks a lot different here.


### 

\footnotesize
```{r}
abortimport %>%
  select(abort_impdum, votetrump, abort_never) %>%
  na.omit %>%
  group_by(abort_impdum) %>% tally() %>%
  mutate(prop = n/sum(n),
         contreff = c(9.04, 28.65),
         product = prop*contreff,
         parteff = sum(product))
```

\normalsize

The partial effect of issue importance is about 20.2 percentage points.

### 

\footnotesize
```{r}
abortimport %>%
  select(abort_impdum, votetrump, abort_never) %>%
  na.omit %>%
  group_by(abort_never) %>% tally() %>%
  mutate(prop = n/sum(n),
         contreff = c(.15, 19.76),
         product = prop*contreff,
         parteff = sum(product))
```

\normalsize

The partial effect of attitudes about abortion is about 3.11 percentage points.


###

```{r abortion-opinions-vote-trump, eval=T, echo=F, fig.width=14, fig.height=8.5, warning=F, message=F}
abortimport %>%
  group_by(abort_impdum, abort_never) %>%
  summarize(mean = mean(votetrump, na.rm=T)) %>%
  na.omit %>% ungroup() %>%
  mutate(abort_impdum = c("Not as Important", "Not as Important",
                          "Very or Extremely Important", "Very or Extremely Important"),
         abort_impdum = fct_inorder(abort_impdum)) %>%
  ggplot(.,aes(x=as.factor(abort_never), y = mean,
               group=abort_impdum,color=abort_impdum,
               linetype=abort_impdum, shape=abort_impdum)) + 
  theme_steve_web() +
  geom_point(size=4) + geom_line(size=1.1) +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(labels = c("At Least\nSometimes Permit", "Never Permit")) +
  scale_color_brewer(palette="Paired") +
  labs(title = "A Line Chart of Abortion Attitudes and the Trump Vote, Controlling for the Importance of Abortion as an Issue",
       subtitle = "The lines show a different effect of abortion attitudes by importance of the issue, suggesting a clear interactive effect.",
       x = "",
       y = "Percentage Saying They Voted for Trump",
       linetype = "", color = "", shape="",
       caption = "Data: American National Election Studies (2020 Exploratory Testing Survey)")
```

<!-- ### Partisanship and Gun Control -->

<!-- ![](tab51.pdf) -->

<!-- ### Zero-order Relationship -->

<!-- Table 5.1 illustrates a **zero-order relationship**. -->

<!-- - This is the effect of *x* on *y* **not** controlling for *z*. -->

<!-- The zero-order effect of partisanship on attitudes toward gun control is about 23 percentage points. -->

<!-- ### Partisanship, Gender, and Gun Control -->

<!-- Consider this arrangement. -->

<!-- ![](tab52.pdf) -->

<!-- ## Controlled and Partial Effects -->
<!-- ### Controlled Comparison Table -->

<!-- Table 5.2 is a **controlled comparison table**. -->

<!-- - It shows the relationship between *x* and *y* for each value of *z*. -->

<!-- These tables reveal two types of relationships. -->

<!-- 1. Controlled effect -->
<!-- 2. Partial effect -->

<!-- ### Controlled Effect -->

<!-- A **controlled effect** is a relationship between *x* and *y* within one value of *z*. -->

<!-- - It could also be between *z* and *y* within one value of *x*. -->

<!-- We obtain the controlled effect of partisanship for both women and men. -->

<!-- - 91.5% of female Dems favor permits to 74.9% of female Republicans. -->
<!-- 	- The controlled effect is 16.6 percentage points. -->
<!-- - 80.8% of male Dems favor permits to 61.1% of male Republicans. -->
<!-- 	- The controlled effect is 19.7 percentage points. -->

<!-- ### Partial Effect -->

<!-- We summarize controlled effects as **partial effect**. -->

<!-- - Summarizes a relationship between two variables taking into account rival variables. -->

<!-- It's tempting, but *don't average the two controlled effects*. -->

<!-- - Doing so assumes the samples are equal. -->

<!-- Instead: weight the controlled effect by percentage of the sample. -->

<!-- - In our case: 16.6 $*$ (.562) + 19.7 $*$ (.438) = 17.96 -->

<!-- The partial effect of partisanship on gun control opinions is 17.96. -->

<!-- ### Partial Effect -->

<!-- What is the partial effect of *gender* on gun control attitudes? -->

<!-- - Sounds weird to ask. No one "increases" in gender. -->

<!-- Follow the **rule of direction for nominal relationships**. -->

<!-- - Treat the left-most column as the base category (here: women). -->

<!-- ### Partial Effect of Gender on Gun Control -->

<!-- Controlled effects: -->

<!-- - 91.5% of female Dems favor permits to 80.8% of male Dems (10.7%). -->
<!-- - 74.9% of female Republicans favor permits to 61.1% of male Republicans (13.8). -->

<!-- Weight the controlled effects to get a partial effect. -->

<!-- - 10.7$*$(.570) + 13.8$*$(.430) = 12.03 -->

<!-- The partial effect of gender on gun control opinions is 12.03. -->

<!-- ## Identifying the Pattern -->
<!-- ### Identifying the Pattern -->

<!-- Ask the following three questions for relationships among *x*, *y*, and *z*. -->

<!-- 1. Does a relationship exist between *x* and *y* in at least one value of *z*? -->
<!-- 2. Is the tendency (i.e. positive or negative) the same at all values of *z*? -->
<!-- 3. Is the magnitude effect the same or close to it in all values of *z*? -->

<!-- ### Identifying the Pattern -->

<!-- If the answer to the first question is no, you can stop there. -->

<!-- - It's a spurious relationship. -->

<!-- If the answer to the second question is no, you can stop there. -->

<!-- - There's an interaction effect. -->

<!-- If the answer to the third question is no, there's an interaction. -->

<!-- - If "yes", it's an additive relationship. -->


<!-- ### Partisanship, Gender, and Gun Control -->

<!-- ![](fig51.pdf) -->


<!-- ## A Clear Example of an Interaction -->
<!-- ### Abortion Opinions, Salience, and Vote Choice -->

<!-- Consider the following controlled comparison table. -->

<!-- ![](tab53.pdf) -->

<!-- ### Abortion Opinions, Salience, and Vote Choice -->

<!-- The controlled effect of abortion opinion: -->

<!-- - 63.2 - 45.9 for "low" = 17.3 -->
<!-- - 82 - 34.9 for "high" = 47.1 -->
<!-- - Partial effect: 17.3 $*$ (.495) + 47.1 $*$ (.505) = 32.3 -->

<!-- Controlled effect of salience: -->

<!-- - 63.2 - 82 for "always permit" = -18.8 -->
<!-- - 45.9 - 34.9 for "not always permit" = 11 -->
<!-- - Partial effect: -18.8 $*$ (.416) + 11 $*$ (.584) = -1.39 -->

<!-- Something already looks a lot different here. -->

<!-- ### Abortion Opinions, Salience, and Vote Choice -->

<!-- ![](fig52.pdf) -->

# Conclusion
### Conclusion

No causal statement can be made as a zero-order relationship

- This will get more complicated in multiple regression.
- Fortunately, computers do the heavy lifting for us.

Get comfortable making these types of controlled comparisons within a simple three-variable context. 