---
title: Post-estimation Simulation
subtitle: POSC 3410  -- Quantitative Methods in Political Science
author: Steven V. Miller
institute: Department of Political Science
titlegraphic: /Dropbox/teaching/clemson-academic.png
date: 
fontsize: 10pt
output:
 beamer_presentation:
    template: ~/Dropbox/miscelanea/svm-r-markdown-templates/svm.latex.beamer.template
#    keep_tex: true
# toc: true
 slide_level: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE)
library(car)
library(arm)
Data <- read.csv("~/Dropbox/teaching/posc3410/post-estimation-simulation/gss14vote.csv", na.strings=c(".a",".d",".i",".n"))
Data$voteobama <- with(Data, recode(pres12, "1=1; 2:4=0"))
Data$voteromney <- with(Data, recode(pres12, "1=0; 2=1; 3:4=0"))
Data$white <- with(Data, recode(race, "1=1; 2:3=0"))
Data$black <- with(Data, recode(race, "1=0; 2=1; 3=0"))
Data$otherrace <- with(Data, recode(race, "1:2=0; 3=1"))
Data$collegeed <- with(Data, recode(degree, "0:2=0; 3:4=1"))
Data$partyid <- with(Data, recode(partyid, "7=NA"))
Data$female <- with(Data, recode(sex, "1=0; 2=1"))
Data$z.age <- arm::rescale(Data$age)
Data$z.partyid <- arm::rescale(Data$partyid)
Data$z.rincom06 <- arm::rescale(Data$rincom06)
library(Zelig)
Data2 <- with(Data, data.frame(voteromney, age, white, black, otherrace, collegeed, partyid, female, rincom06))
Data2 <- na.omit(Data2)

M3 <- zelig(voteromney ~ age + female + collegeed + white + 
              black + partyid + rincom06, 
model = "logit", data = Data2)

M3.bsdem <- setx(M3, black = 1, white = 0, partyid = 0)
M3.bsrep <- setx(M3, black = 1, white = 0, partyid = 6)

M3.sim <- sim(M3, x = M3.bsdem, x1 = M3.bsrep)

M3.bsdemtosrep <- setx(M3, black = 1, white = 0, partyid = 0:6)

M3.sim2 <- sim(M3, x = M3.bsdemtosrep)
```

# Introduction
### Goal for Today

*Provide intuitive quantities of interest from your regression.*

### Readable Regression Tables

Remember: your analysis should be as easily interpretable as possible.

- I should get a preliminary glimpse of effect size from a regression.
- Your *y*-intercept should be meaningful.

Standardizing variables helps.

- Creates meaningful zeroes (i.e. the mean).
- Coefficients communicate magnitude changes in *x*.
- Standardizing by two SDs allows for easy comparison with binary predictors.

### Satisfy Your Audience

You need to relate your analysis to both me and your grandma.

- I will obviously know/care more about technical details.
- Grandma may not, but she may be a more important audience than me.

Her inquiries are likely be understandable. Examples:

- What's the expected tolerance of abortion for an 18-year-old black man?
- What's the increased probability of voting for a Republican for an increase of $20k in yearly income?

These are perfectly reasonable questions to ask of your analysis.

- If your presentation isn't prepared to answer her questions, you're not doing your job.

### Statistical Presentations

Statistical presentations should:

1. Convey precise estimates of quantities of interest.
2. Include reasonable estimates of *uncertainty* around those estimates.
3. Require little specialized knowledge to understand Nos. 1 and 2.
4. Not bombard the audience with superfluous information.

We will do this with post-estimation simulation using draws from a multivariate normal distribution (King et al. 2000).

# Estimating Uncertainty with Simulation
## Systematic and Stochastic Components
### Estimating Uncertainty with Simulation

Any statistical model has a stochastic and systematic component.

- **Stochastic**: $Y_i \sim f(y_i \thinspace | \thinspace \theta_i, \alpha )$
- **Systematic**:  $\theta_i = g(x_i, \beta)$

For a simple OLS model (i.e. a linear regression):

\begin{eqnarray}
Y_i &=& N(\mu_i, \thinspace \sigma^2)    \nonumber \\
\mu_i &=&  X_i\beta  \nonumber
\end{eqnarray}

### Understanding our Uncertainty

We have two types of uncertainty.

1. **Estimation uncertainty**
	-  Represents systematic components; can be reduced by increasing sample size.
2. **Fundamental uncertainty**
	- Represents stochastic component; exists no matter what (but can be modeled).

### Getting our Parameter Vector

We want a **simulated parameter vector**, denoted as:

$$
\hat{\gamma} \sim vec( \hat{\beta}, \hat{\alpha})
$$

Central limit theorem says with a large enough sample and bounded variance:

$$
\tilde{\gamma} \sim N( \hat{\gamma} , \hat{V}(\hat{\gamma}))
$$

In other words: distribution of quantities of interest will follow a multivariate normal distribution with mean equal to $\hat{\gamma}$, the simulated parameter vector.

### Getting our Quantities of Interest

This is a mouthful! Let's break the process down step-by-step.

1. Run your regression. Get your results.
2. Choose values of explanatory variable (as you see fit).
3. Obtain simulated parameter vector from estimating systematic component.
4. Simulate the outcome by taking random draw from the stochastic component.

Do this *m* times (typically *m* = 1000) to estimate full probability distribution of $Y_c$.

- Expected value $E(Y_c)$ = predicted value for linear models. It just averages over the fundamental uncertainty.

# An Application with Vote Choice in 2012 (2014 GSS)
### An Application with Zelig

Don't worry! We have software to make this easier.

- We'll be using the Zelig package in R.

I'll also be using sample data from my Github page.

- Question: What explains who voted for Romney or Obama in the 2012 presidential election?

### Understanding our Sample Data

I took the 2014 wave of GSS data on American public opinion.

- **voteromney**: 1 if respondent (*r*) voted for Romney; 0 if *r* voted for someone else.
	- I also have an Obama version of this variable.
- **age**: age of respondent in years (18:89).
	- Note: Regrettably, GSS stops coding year after 89+.
- **race**: three category nominal variable (white, black, other).
- **collegeed**: 1 if *r* has at least a bachelors; 0 if otherwise.
- **partyid**: strength of affiliation with Republican party (0:6).
	- 0 = strong Democrat; 6 = strong Republican
- **female**: 1 if *r* is a woman; 0 if *r* is a man.
- **rincom06**: Income categories (1:25).
	- 1 = under $1,000 earned in 2013; 25 = $150k or over.

### A Sample Regression

\footnotesize
```{r}
M1 <- glm(voteromney ~ age + female + collegeed + white + black + partyid + rincom06, 
          data=Data, family=binomial(link="logit"))
summary(M1)

```

\normalsize

### Standardizing the Variables

\footnotesize
```{r}
M2 <- glm(voteromney ~ z.age + female + collegeed + white + black + 
            z.partyid + z.rincom06, data=Data, family=binomial(link="logit"))
summary(M2)

```

\normalsize

### Something Nana Might Ask

What's the effect of Republican party ID on black voters?

- Let's see!

First, we re-estimate the model in Zelig.

- We also have to subset for complete cases.
- Zelig doesn't play nice with missing data.

### The Model in Zelig

```{r eval=FALSE}
M3 <- zelig(voteromney ~ age + female + collegeed + white + 
              black + partyid + rincom06, 
model = "logit", data = Data2)
```

Notice Zelig's post-estimation simulation doesn't require intuitive zeroes. 

- You should still do it for a regression table, though.
- Again, not having intuitive zeroes can still break a more complicated model.

### Answering Nana's Question

What's the probability of a black vote for Romney among strong Republicans? How about strong Democrats?


```{r eval=FALSE}
M3.bsdem <- setx(M3, black = 1, white = 0, partyid = 0)
M3.bsrep <- setx(M3, black = 1, white = 0, partyid = 6)

M3.sim <- sim(M3, x = M3.bsdem, x1 = M3.bsrep)
summary(M3.sim)

```


### Answering Nana's Question

\tiny
```{r}
summary(M3.sim)
```

\normalsize

### Another Question from Nana

What's the effect of Republican party identification on Black voters?

```{r eval=FALSE}
M3.bsdemtosrep <- setx(M3, black = 1, white = 0, partyid = 0:6)

M3.sim2 <- sim(M3, x = M3.bsdemtosrep)

plot.ci(M3.sim2, xlab = "Party Identification (D to R)",
        ylab = "E(Vote for Romney)",
        main = "The Effect of Republican Party Identification 
        on \n a Black Person's Vote for Romney",
        ci=c(90,95)
)
```

### A Pretty Graph

```{r echo=FALSE}
plot.ci(M3.sim2, xlab = "Party Identification (D to R)",
        ylab = "E(Vote for Romney)",
        main = "The Effect of Republican Party Identification on \n a Black Person's Vote for Romney",
        ci=c(90,95)
)
```

### What's Going on With the Black Vote among the Republicans?

Notice the huge confidence intervals for Black Republicans? What do you think is happening there?

### There Aren't Many Black Republicans

```{r}
table(Data$black, Data$partyid)
```

Note: Black respondents are the row; partyid is the column

-  Also: 4 = "independent, near Republican"; 5 = "not a strong Republican"; 6 = "strong Republican"

### There Weren't Many Black Voters for Romney Either

```{r}
table(Data$black, Data$voteromney)
```

Note: Black respondents are the row; Romney votes are the column.

- If you're curious, two of those three black votes for Romney came from self-identified strong Republicans.
- One actually came from a self-identified strong Democrat!

The relative dearth of black Republicans and black Romney voters adds to the uncertainty in our predictions.

# Conclusion
### Conclusion

Regression provides all-else-equal effect sizes across the range of the data.

- You can extract meaningful quantities of interest from regression output itself.
- Typically, you'll need more to answer substantive questions and provide meaningful quantities of interest.

Post-estimation simulation from a multivariate normal distribution does this.

- When you start doing this yourselves, be prepared to provide quantities of interest for your audience.
