---
title: "Central Tendency and Dispersion"
subtitle: POSC 3410  -- Quantitative Methods in Political Science
author: Steven V. Miller
institute: Department of Political Science
titlegraphic: /Dropbox/teaching/clemson-academic.png
date: 
fontsize: 10pt
output:
 beamer_presentation:
    template: ~/Dropbox/miscelanea/svm-r-markdown-templates/svm-latex-beamer.tex
    latex_engine: xelatex
    dev: cairo_pdf
    fig_caption: false
    slide_level: 3
make149: true
mainfont: "Open Sans"
titlefont: "Titillium Web"
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
---

```{r setup, include=FALSE, cache=F, message=F, warning=F, results="hide"}
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_chunk$set(fig.path='figs/')
knitr::opts_chunk$set(cache.path='cache/')

knitr::opts_chunk$set(
                  fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      }
                  )
```

```{r loadstuff, include=FALSE}
library(tidyverse)
library(stevemisc)
library(post8000r)
library(qs)
library(knitr)
library(kableExtra)

# USA <- qread("~/Dropbox/data/wvs/wvs6wave-20180912.qs") %>% rename_all(tolower) %>%
#   filter(s003 == 840)
# 
# USA %>% filter(s002 == 5) %>%
#     select(s020, x047cs) %>%
#     mutate(inccat = x047cs - 840040) -> USA2006

GSS <- qread("/home/steve/Dropbox/data/gss/GSS_spss-2018/gss7218.qs") %>% select(year, region, attend, tvhours)
```

# Introduction
### Goal for Today

*Describe variables by reference to central tendency and dispersion.*

### Defining and Measuring Variables

Last lecture focused on a typology of variables.

1. Nominal
2. Ordinal
3. Interval

Correct classification will condition how we can *describe* variables.

# Central Tendency
### Central Tendency

The most common description of interest is the **central tendency**.

- This is the variables "typical", or "average" value.
- This takes on different forms contingent on variable type.

Think of what follows as a "tool kit" for researchers.

- More precise variables allow for more precise measures.
- Use the right tool for the job, if you will.

### Mode

The **mode** is the most basic central tendency statistic.

- It identifies the most frequently occurring value.

Suppose I have a random sample of 50 students and measured party affiliation.

- Democrats: 26; Republicans: 20; Others: 4

What's the modal category?

### Mode

If I randomly grabbed a student from that sample and guessed "Democrat", I would be right 26 times of 50 (on average).

- No other guess, on average, would be as good.

This is the only central tendency statistic for nominal variables.

### Median

The **median** is the middlemost value.

- It's the most precise statistic for ordinal variables.
- It's a useful robustness check for interval variables too.

Formally, a median *m* exists when the following equalities are satisfied.

\begin{equation}
	P(X \le m) \ge \frac{1}{2} \; \textrm{and} \;  P(X \ge m) \ge \frac{1}{2}
\end{equation}

### Finding the Median

Order the observations from lowest to highest and find what value lies in the exact middle.

- The median is the point where half the values lie below and half are above.
- We can do this when our variables have some kind of "order".
- Medians of nominal variables are nonsensical.

### Mean

The arithmetic **mean** is used only for interval variables.

- This is to what we refer when we say "average".

Formally, *i* through *n*:

\begin{equation}
	\frac{1}{n}\Sigma x_i
\end{equation}

We can always describe interval variables with mode and median.

- We cannot do the same for ordinal or nominal with the mean.

# Dispersion
### Dispersion

We also need to know variables by reference to its **dispersion**.

- i.e. "how average is 'average'?"
- How far do variables deviate from the typical value?
- If they do, measures of central tendency can be misleading.

The interval variable with no dispersion problem is one in which the mode, median, and mean are the same value.

- This will not happen when there is a significant **skew**, or a **bimodal** distribution.

### Frequency Distribution

A **frequency distribution** is a summary of a variable's values.

 <!-- ![](tab22.pdf) -->
 
```{r, echo=F, message=F, warning=F, results="asis"}
skable <- function(data, caption) {
  knitr::kable(data, caption = caption, format = "latex", booktabs = TRUE, longtable =T, linesep = "") %>%
    kable_styling(full_width = F, 
                  latex_options = c("repeat_header", "hold_position")) %>%
    row_spec(0, bold= TRUE)
}

GSS %>% select(year, region) %>%
  filter(year == 2018) %>%
  mutate(division = as.character(str_to_title(as_factor(region)))) %>%
  select(-region) %>%
  mutate(division = dplyr::recode(division,
                                "E. Nor. Central" = "East North Central",
                                "W. Sou. Central" = "West South Central",
                                "W. Nor. Central" = "West North Central",
                                "E. Sou. Central" = "East South Central")) %>%
  left_join(., States %>% select(division, region)) %>%
  group_by(region) %>% tally() %>%
  mutate(region = ifelse(region == "North Central", "Midwest", region),
         perc = n/sum(n),
         perc = paste0(mround2(perc),"%")) %>%
  bind_rows(., tibble(region = "Total", perc = "100%" )) %>%
  mutate(holdme = sum(n,na.rm=T),
         n = ifelse(is.na(n), holdme, n)) %>% select(-holdme) %>%
  kable(., format="latex",
        caption = "Region of Residence (General Social Survey, 2018)",
        col.names = c("Region","Frequency","Percentage"),
        align = c("l","c","c"), booktabs=TRUE, longtable=TRUE) %>%
  row_spec(5, italic=TRUE) %>%
  row_spec(0, bold=TRUE) 

```

### Cumulative Percentage

A **cumulative percentage** is the percentage of cases at or below a given value.

```{r, echo=F, message=F, warning=F, results="asis"}

GSS %>%
  filter(year == 2018) %>%
  arrange(attend) %>%
  mutate(attend_f = str_to_title(as_factor(attend))) %>%
  filter(!is.na(attend)) %>%
  mutate(attend_f = forcats::fct_inorder(attend_f)) %>%
  group_by(attend_f) %>% tally() %>%
  mutate(attend_f = c("Never or Less Than Once a Year",
                      "Never or Less Than Once a Year",
                      "Once a Year",
                      "Several Times a Year",
                      "Once a Month",
                      "2-3 Times a Month",
                      "Nearly Every Week",
                      "Every Week or More",
                      "Every Week or More")) %>%
  group_by(attend_f) %>% mutate(nt = sum(n)) %>%
  distinct(attend_f, nt) %>%
  ungroup() %>%
  mutate(perc = nt/sum(nt),
         cumperc = cumsum(perc)) %>%
  mutate(perc = paste0(mround2(perc), "%"),
         cumperc =paste0(mround2(cumperc), "%")) %>%
   kable(., format="latex",
        caption = "Attendance at Religious Services (General Social Survey, 2018)",
        col.names = c("Attendance","Frequency","Percentage", "Cumulative Percentage"),
        align = c("l","c","c","c"), booktabs=TRUE, longtable=TRUE,linesep='') %>%
  # row_spec(5, italic=TRUE) %>%
  row_spec(0, bold=TRUE) 
  

```

###

```{r church-attendance-over-time, eval=T, echo=F, fig.width=14, fig.height=8.5, warning=F, message=F}

GSS %>%
  mutate(attend_f = str_to_title(as_factor(attend))) %>% 
  filter(!is.na(attend_f)) %>% #distinct(attend_f) %>%
  mutate(attend_f = dplyr::recode(attend_f,
                                  "Never" = "Never or Less Than Once a Year",
                                  "Once A Year" = "Once or Several Times a Year",
                                  "Once A Month" = "At Least Once a Month",
                                  "Lt Once A Year" = "Never or Less Than Once a Year",
                                  "Sevrl Times A Yr" = "Once or Several Times a Year",
                                  "More Thn Once Wk" = "Every Week or More",
                                  "Every Week" = "Every Week or More",
                                  "2-3x A Month" = "At Least Once a Month",
                                  "Nrly Every Week" = "At Least Once a Month")) %>%
  mutate(attend_f = forcats::fct_relevel(attend_f,
                                         "Never or Less Than Once a Year",
                                         #"Once a Year",
                                         "Once or Several Times a Year",
                                     #    "Once a Month",
                                         "At Least Once a Month",
                                      #   "Nearly Every Week",
                                         "Every Week or More")) %>%
  group_by(year, attend_f) %>%
  tally() %>%
  group_by(year) %>%
  mutate(perc = n/sum(n)) %>% #filter(attend_f %nin% c("Once a Month", "2-3 Times a Month")) %>%
  ggplot(.,aes(year, perc, color=attend_f,linetype=attend_f)) + geom_line(size=1.1) +
  theme_steve_web()  +  scale_color_brewer(palette = "PuOr") +
  scale_x_continuous(breaks = seq(1970,2020, by=4)) +
  labs(color="",linetype="",
       title = "The Variation of Self-Reported Church Attendance in the United States, 1972-2018",
       x = "",
       y = "Percentage of Responses in a Given Survey Year",
       caption = "Data: General Social Survey, 1972-2018",
       subtitle = "In these truncated categories, the ''never'' group went from being the smallest in 1972 to the clear largest in 2018.")

```

###

```{r bar-chart-region, eval=T, echo=F, fig.width=14, fig.height=8.5, warning=F, message=F}
GSS %>% select(year, region) %>%
  filter(year == 2018) %>%
  mutate(division = as.character(str_to_title(as_factor(region)))) %>%
  select(-region) %>%
  mutate(division = dplyr::recode(division,
                                "E. Nor. Central" = "East North Central",
                                "W. Sou. Central" = "West South Central",
                                "W. Nor. Central" = "West North Central",
                                "E. Sou. Central" = "East South Central")) %>%
  left_join(., States %>% select(division, region)) %>%
  group_by(region) %>% tally() %>%
  mutate(region = ifelse(region == "North Central", "Midwest", region),
         perc = n/sum(n),
         perc = paste0(mround2(perc),"%")) %>%
  ggplot(.,aes(region, n)) + 
  theme_steve_web() +
  geom_bar(stat="identity", color="black", fill="#619cff") +
   geom_text(aes(label=n), vjust=-.5, colour="black",
            position=position_dodge(.9), size=4, family = "Open Sans") +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Condensed Census Region",
       y = "Number of Observations",
       title = "A Bar Chart of Region of Residence in the General Social Survey (2018)",
       subtitle = "Your mileage may vary, but I think there's more value in bar charts for basic descriptive stuff.",
       caption = "Data: General Social Survey, 2018.")
```

###

```{r bar-chart-churchatd, eval=T, echo=F, fig.width=14, fig.height=8.5, warning=F, message=F}
GSS %>%
  filter(year == 2018) %>%
  arrange(attend) %>%
  mutate(attend_f = str_to_title(as_factor(attend))) %>%
  filter(!is.na(attend)) %>%
  mutate(attend_f = forcats::fct_inorder(attend_f)) %>%
  group_by(attend_f) %>% tally() %>%
  mutate(attend_f = c("Never or Less Than Once a Year",
                      "Never or Less Than Once a Year",
                      "Once a Year",
                      "Several Times a Year",
                      "Once a Month",
                      "2-3 Times a Month",
                      "Nearly Every Week",
                      "Every Week or More",
                      "Every Week or More")) %>%
  group_by(attend_f) %>% mutate(nt = sum(n)) %>%
  distinct(attend_f, nt) %>%
  ungroup() %>%
  mutate(perc = nt/sum(nt),
         cumperc = cumsum(perc)) %>%
  mutate(attend_f = fct_inorder(attend_f)) %>%
  ggplot(.,aes(attend_f, nt)) + 
  theme_steve_web() +
  geom_bar(stat="identity", color="black", fill="#619cff") +
   geom_text(aes(label=nt), vjust=-.5, colour="black",
            position=position_dodge(.9), size=4, family = "Open Sans") +
  labs(x = "Self-Reported Church Attendance",
       y = "Number of Observations",
       caption = "Data: General Social Survey, 2018.",
       title = "A Bar Chart of Self-Reported Church Attendance in the General Social Survey (2018)",
       subtitle = "A simple bar chart helps less with cumulative percentages, but it'll better point you in the direction of potential bimodality.")
```


###

```{r bar-chart-tvwatching, eval=T, echo=F, fig.width=14, fig.height=8.5, warning=F, message=F}

GSS %>%
  filter(year == 2018) %>%
  summarize(meantv = mean(tvhours,na.rm=T),
            mediantv = median(tvhours, na.rm=T)) -> sumssums


GSS %>%
  filter(year == 2018) %>%
  group_by(tvhours) %>% tally() %>% na.omit %>%
  ggplot(.,aes(tvhours, n)) +
  theme_steve_web() +
  geom_bar(stat="identity", color="black", fill="#619cff") +
    geom_vline(xintercept = sumssums$meantv, linetype="dashed") +
  geom_vline(xintercept = sumssums$mediantv, linetype="dotted") +
  scale_x_continuous(breaks = 0:24) +
     geom_text(aes(label=n), vjust=-.5, colour="black",
            position=position_dodge(.9), size=3.5, family = "Open Sans") +
  labs(title = "Self-Reported Hours Watching TV (General Social Survey, 2018)",
       subtitle = "A simple bar chart will also help visualize skew in an intuitive way.",
       x = "Self-Reported Hours Watching TV",
       y = "Number of Observations",
       caption = "Data: General Social Survey, 2018. No one is watching 24 hours of TV. You're not fooling anyone.\nMode and median (2) captured in dotted line. Mean in dashed line.")



```


# Conclusion
### Conclusion

Here are some final thoughts.

- There is a reason we discuss "median income" and not the "average income".
- The mean of a dummy variable communicates the percentage of 1s, divided by 100.
- Skew is mostly a problem of interval variables, and a problem of degree.

Always look carefully at your data!
